### 2023-09-26 chapters/paper_hierarchical.tex
Move table captions to top

\section{OOD detection with hierarchical VAEs}
\subsection{A bound for semantic OOD detection}
If the lowest latent variable in the VAE hierarchy codes for a large part of the low-level features required to reconstruct the input with high accuracy, as exemplified in \cref{fig_hierarchical:reconstructions-fashionmnist}-\ref{fig_hierarchical:biva-reconstructions-celeba}, then $p_\thetab(\xb|\zb_1)$ will be high for both in- and out-of-distribution data.
Hence, any OOD detection capabilities based on the ELBO $\mathcal{L} = \mathbb{E}_{q_\phib(\zb|\xb)}[\log p_\thetab(\xb|\zb_1)] - D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel  p(\zb))$ from \cref{eq_hierarchical:elbo} relies on the KL-term for OOD detection. For a bottom-up hierarchical VAE, the KL-term $D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel  p(\zb))$ can be expressed by a hierarchical sum% over the hierarchy
% \begin{multline}
%     \mathcal{L}(\xb) =  \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \log p_\thetab(\xb|\zb_1) \\
%                      + \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb_i|\zb_{i+1})}{q_\phib(\zb_i|\zb_{i-1})} 
%                      + \log \frac{p_\thetab(\zb_L)}{q_\phib(\zb_L|\zb_{L-1})} \Big] .
% \end{multline}
% \begin{multline}
%     D_{\mathrm{KL}}( q(\zb|\xb) \parallel p(\zb)) = \\
%     \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb_i|\zb_{i+1})}{q_\phib(\zb_i|\zb_{i-1})} + \log \frac{p_\thetab(\zb_L)}{q_\phib(\zb_L|\zb_{L-1})} \Big] .
% \end{multline}
\begin{equation}
    \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb_i|\zb_{i+1})}{q_\phib(\zb_i|\zb_{i-1})} + \log \frac{p_\thetab(\zb_L)}{q_\phib(\zb_L|\zb_{L-1})} \Big] \ .
\end{equation}
In general, the absolute log-ratios grow with $\mathrm{dim}(\zb_i)$ as the individual log probability terms are computed by summing over the dimensionality of $\zb_i$.
This means that the value of the KL-term is dominated by terms where $\zb_i$ is high-dimensional. We refer to Appendix C for a more detailed argument.
Since hierarchical VAEs are generally constructed with a bottleneck type structure, the terms corresponding to latent variables towards the top of the hierarchy will have a vanishing influence on the value of the KL-term.
However, as the semantic information most relevant for OOD detection has a tendency to be represented in the top-most latent variables, this makes OOD detection using the regular ELBO difficult, even for state-of-the-art models.
This behavior has also been reported by \textcite{xiao_likelihood_2020}.

To shift the ELBO from primarily being based on the approximate posterior of the lowest latent variables to instead focus on the conditional prior, \textcite{maaloe_biva_2019} introduced slightly different likelihood lower bound defined as
\begin{equation}\label{eq_hierarchical:biva->k}
    \mathcal{L}^{>k} = \mathbb{E}_{p_\thetab(\zb_{\leq k}|\zb_{>k}) q_\phib(\zb_{>k}|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p_\thetab(\zb_{>k})}{q_\phib(\zb_{>k}|\xb)} \right]
\end{equation}
where $k\in\{0,1,\dots,L\}$ (see Appendix for the derivation).
We note that $\mathcal{L}^{>0}$ is the regular ELBO (\cref{eq_hierarchical:elbo}) and that empirically we always observe that $\mathcal{L}\geq\mathcal{L}^{>k} \, \forall \, k$ although this need not hold in general.
The core idea behind this variation on the ELBO is to sample the $k$ lowest latent variables from the conditional prior $\zb_1,\dots,\zb_l \sim p_\thetab(\zb_{\leq k}|\zb_{>k})$ and only the $L-k$ highest from the approximate posterior $\zb_{k+1},\dots,\zb_L \sim q_\phib(\zb_{>k}|\xb)$.
Importantly, this has the effect that the data likelihood $p(\xb|\zb)$ is dependent on the approximate posterior through a latent variable $\zb_{k+1}$ different from $\zb_1$ for all $k \geq 1$.
Thereby, the likelihood can be evaluated with a reconstruction from each of the latent variables $\zb_k$ of the hierarchical VAE.
Hence, we can now test how well the input $\xb$ is reconstructed from each latent variable.
The notation $\mathcal{L}^{>k}$ highlights that for latent variables $\zb_{>k}$, the bound is the regular ELBO while for the latent variables $\zb_{\leq k}$, the bound is evaluated using the (conditional) prior rather than the approximate posterior as the proposal distribution.

@@

The OOD scores considered in this research all assume that what discriminates an out-of-distribution from an in-distribution data point are semantic, high-level features. Clearly, if this is not the case and the difference instead lies in low-level statistics, the scores would likely fail. We hypothesize that a complementary bound to \cref{eq_hierarchical:biva->k}, $\mathcal{L}^{<l}$ described in Appendix E, might be useful in these cases, but leave further examination to future work.

@@

\paragraph{Tasks} We follow existing literature \cite{nalisnick_deep_2019, hendrycks_deep_2019} and evaluate our method by setting up OOD detection tasks from FashionMNIST \cite{xiao_fashionmnist_2017} to MNIST \cite{lecun_gradientbased_1998} and from CIFAR10 \cite{krizhevsky_learning_2009} to SVHN \cite{netzer_reading_2011}.
For each experiment we train our model on the train split of the former dataset and test its ability to recognize the test split of the latter dataset as OOD from the test split of the former dataset.
We use the standard train/test splits for the datasets.
More details on the datasets can be found in the Appendix.

@@

\paragraph{Models} For each OOD task, we train a simple bottom-up hierarchical VAE with $L$ stochastic layers which we will refer to as ``HVAE''.
To alleviate posterior collapse we include skip-connections that connect $\zb_i$ to $\zb_{i+2}$ for $i\in\{0, L-2\}$ and $\zb_0\equiv\xb$ in both the inference and generative models \cite{dieng_avoiding_2019} and employ the \textit{free bits} scheme with $\lambda=2$ \cite{kingma_improved_2016}.
We use weight-normalization \cite{salimans_weight_2016} on all weights and residual networks in the deterministic paths. 
A graphical representation of this model can be seen in \cref{fig_hierarchical:hvae-graphical-model}.
We use a Bernoulli output distribution for FashionMNIST/MNIST and a discretized mixture of logistics output distribution \cite{salimans_pixelcnn_2017} for CIFAR10/SVHN.
We use $L=3$ for grey-scale images and $L=4$ for natural images.
% For CIFAR/SVHN, we also train a BIVA model \cite{maaloe_biva_2019} with $L=10$ and similar configuration as used by the original paper\footnote{Code available at \url{github.com/larsmaaloee/BIVA} and \url{github.com/vlievin/biva-pytorch}}.
Full model details are in the Appendix.

@@

The likelihoods for our trained models are in \cref{tab_hierarchical:bits-per-dim-ood} alongside baseline results for in-distribution and OOD data.
The main results of the paper on the OOD tasks can be seen along with comparisons to the baseline methods in \cref{tab_hierarchical:rocauc-ood}.
We note that for all our results, the value of the score ($\mathcal{L}^{>k}$ and $LLR^{>k}$) for the training and test splits of the in-distribution data was observed to have the same empirical distribution to within sampling error hence yielding an AUROC score of $\approx0.5$ as expected.
Results on additional commonly used datasets are found in Appendix G.

\begin{table}[t!]
    \centering
    \resizebox{0.8\columnwidth}{!}{%
    \begin{tabular}{lrrrrr}
        \toprule
         Method & Dataset & \multicolumn{4}{c}{Avg. bits/dim}\\
        %   & & $\log p(x)$ & $\mathcal{L}^{>k_1}$ & $\mathcal{L}^{>k_2}$ & $\mathcal{L}^{>k_3}$\\
          & & $\log p(x)$ & $\mathcal{L}^{>1}$ & $\mathcal{L}^{>2}$ & $\mathcal{L}^{>3}$\\
         \midrule
         \multicolumn{6}{c}{\textbf{Trained on FashionMNIST}} \\
         \midrule
         \multirow{2}{*}{Glow}
            & FashionMNIST & 2.96 & - & - & \\
            & MNIST & 1.83 & - & - & \\
         \multirow{2}{*}{HVAE (Ours)}
            & FashionMNIST & 0.420 & 0.476 & 0.579 & - \\
            & MNIST & 0.317 & 0.601 & 0.881 & - \\
         \midrule
         \multicolumn{6}{c}{\textbf{Trained on CIFAR10}} \\
         \midrule
         \multirow{2}{*}{Glow}
          & CIFAR10 & 3.46 & - & - & \\
          & SVHN & 2.39 & - & - & \\
         \multirow{2}{*}{HVAE (Ours)}
            & CIFAR10 & 3.74 & 17.8 & 54.3 & 75.7 \\  % log p(x) = 8010.03
            & SVHN & 2.62 & 10.2 & 64.0 & 93.9 \\
         \multirow{2}{*}{BIVA (Ours)}
          & CIFAR10 & 3.46 & 8.74 & 19.7 & 37.3 \\
          & SVHN & 2.35 & 6.62 & 25.1 & 59.0 \\
         \bottomrule
    \end{tabular}
    }
    \caption[Average bits per dimension of different datasets for generative models trained on FashionMNIST and CIFAR10.]{%
    Average bits per dimension of different datasets for models trained on FashionMNIST and CIFAR10.
    For the hierarchical models we include the $\mathcal{L}^{>k}$ bounds.
    The likelihoods of training and test splits of the in-distribution data are all cases close.
    Since we train on dynamically binarized FashionMNIST, our bits/dim are smaller than for Glow.
    As $k$ is increased for the $L^{>k}$ bound, the bound gets looser but the model eventually assigns higher likelihood to the in distribution data than to the OOD data.
    Glow refers to \textcite{kingma_glow_2018, nalisnick_deep_2019}.
    BIVA refers to our implementation of \textcite{maaloe_biva_2019}.}
    \label{tab_hierarchical:bits-per-dim-ood}
    \vspace{0mm}
\end{table}

@@

\subsection{Likelihood-ratio-based OOD detection}
\begin{figure}[t]
    \centering
    \begin{subfigure}[l]{0.495\columnwidth}
        \includegraphics[width=1\columnwidth]{paper_hierarchical/roc-FashionMNIST-test-MNIST-test-ll-and-llr-IW250_sohau.pdf}
        \caption{}
        % \caption{ROC curves with AUROC score for detecting MNIST as OOD with the HVAE model trained on FashionMNIST.
        % A ROC curve is plotted for each of the $\mathcal{L}^{>k}$ bounds including the ELBO along with one for the best-performing log likelihood-ratio $LLR^{>1}$.}
        \label{fig_hierarchical:FMNIST-roc-llr}
    \end{subfigure}
    \hfill
    \begin{subfigure}[r]{0.495\columnwidth}
        \includegraphics[width=1\columnwidth]{paper_hierarchical/roc-biva-CIFAR10-SVHN-ll-and-llr_sohau.pdf}
        \caption{}
        % \caption{ROC curves with AUROC score for detecting SVHN as OOD with the BIVA model trained on CIFAR10.
        % A ROC curve is plotted for each of the $\mathcal{L}^{>k}$ bounds including the ELBO along with one for the best-performing log likelihood-ratio $LLR^{>2}$.}
        \label{fig_hierarchical:CIFAR10-roc-llr}
    \end{subfigure}
    \caption[ROC curves for out of distribution detection (MNIST/FashionMNIST and SVHN/CIFAR10).]{%
        ROC curves with AUROC score for detecting \subref{fig_hierarchical:FMNIST-roc-llr} MNIST as OOD with the HVAE model trained on FashionMNIST and \subref{fig_hierarchical:CIFAR10-roc-llr} SVHN as OOD with the BIVA model trained on CIFAR10. 
        A ROC curve is plotted for each of the $\mathcal{L}^{>k}$ bounds including the ELBO along with one for the best-performing log likelihood-ratio $LLR^{>1}$.
    }
\end{figure}
We now move to the likelihood ratio-based score.
We find that $LLR^{>k}$ separates the OOD MNIST data from in-distribution FashionMNIST to a higher degree than the likelihood estimates as can be seen by the empirical densities of the score in \cref{fig_hierarchical:FMNIST-llr}.
We note that the likelihood ratio between the ELBO and the $\mathcal{L}^{>k}$ bound provides the highest degree of separation of MNIST and FashionMNIST as measured by the AUROC$\uparrow$ for $k=1$ smaller than $L$.
This is not surprising since the value of $k$ that provides the maximal separation to the reference in-distribution dataset need not be the one for which $\mathcal{LLR}^{>k}$ is overall maximal for the OOD dataset.
We also visualize the ROC curves resulting from using the $LLR^{>k}$ score for OOD detection on both FashionMNIST/MNIST and CIFAR10/SVHN and compare it to the ROC curves resulting from the different $\mathcal{L}^{>k}$ bounds in Figures \ref{fig_hierarchical:FMNIST-roc-llr} and \ref{fig_hierarchical:CIFAR10-roc-llr}, respectively.
On both datasets we see significantly better discriminatory performance when using the $LLR^{>k}$ score.

>>>>

\section{OOD detection with hierarchical VAEs}
\subsection{A bound for semantic OOD detection}
If the lowest latent variable in the VAE hierarchy codes for a large part of the low-level features required to reconstruct the input with high accuracy, as exemplified in \crefrange{fig_hierarchical:reconstructions-fashionmnist}{fig_hierarchical:biva-reconstructions-celeba}, then $p_\thetab(\xb|\zb_1)$ will be high for both in- and out-of-distribution data.
Hence, any OOD detection capabilities based on the ELBO $\mathcal{L} = \mathbb{E}_{q_\phib(\zb|\xb)}[\log p_\thetab(\xb|\zb_1)] - D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel  p(\zb))$ from \cref{eq_hierarchical:elbo} relies on the KL-term for OOD detection. For a bottom-up hierarchical VAE, the KL-term $D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel  p(\zb))$ can be expressed by a hierarchical sum% over the hierarchy
% \begin{multline}
%     \mathcal{L}(\xb) =  \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \log p_\thetab(\xb|\zb_1) \\
%                      + \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb_i|\zb_{i+1})}{q_\phib(\zb_i|\zb_{i-1})} 
%                      + \log \frac{p_\thetab(\zb_L)}{q_\phib(\zb_L|\zb_{L-1})} \Big] .
% \end{multline}
% \begin{multline}
%     D_{\mathrm{KL}}( q(\zb|\xb) \parallel p(\zb)) = \\
%     \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb_i|\zb_{i+1})}{q_\phib(\zb_i|\zb_{i-1})} + \log \frac{p_\thetab(\zb_L)}{q_\phib(\zb_L|\zb_{L-1})} \Big] .
% \end{multline}
\begin{equation}
    \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb_i|\zb_{i+1})}{q_\phib(\zb_i|\zb_{i-1})} + \log \frac{p_\thetab(\zb_L)}{q_\phib(\zb_L|\zb_{L-1})} \Big] \ .
\end{equation}
In general, the absolute log-ratios grow with $\mathrm{dim}(\zb_i)$ as the individual log probability terms are computed by summing over the dimensionality of $\zb_i$.
This means that the value of the KL-term is dominated by terms where $\zb_i$ is high-dimensional. We refer to \cref{sec_hierarchical:analysis} for a more detailed argument.
Since hierarchical VAEs are generally constructed with a bottleneck type structure, the terms corresponding to latent variables towards the top of the hierarchy will have a vanishing influence on the value of the KL-term.
However, as the semantic information most relevant for OOD detection has a tendency to be represented in the top-most latent variables, this makes OOD detection using the regular ELBO difficult, even for state-of-the-art models.
This behavior has also been reported by \textcite{xiao_likelihood_2020}.

To shift the ELBO from primarily being based on the approximate posterior of the lowest latent variables to instead focus on the conditional prior, \textcite{maaloe_biva_2019} introduced slightly different likelihood lower bound defined as
\begin{equation}\label{eq_hierarchical:biva->k}
    \mathcal{L}^{>k} = \mathbb{E}_{p_\thetab(\zb_{\leq k}|\zb_{>k}) q_\phib(\zb_{>k}|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p_\thetab(\zb_{>k})}{q_\phib(\zb_{>k}|\xb)} \right]
\end{equation}
where $k\in\{0,1,\dots,L\}$ (see \cref{sec_hierarchical:derivation_L_geq_k} for the derivation).
We note that $\mathcal{L}^{>0}$ is the regular ELBO (\cref{eq_hierarchical:elbo}) and that empirically we always observe that $\mathcal{L}\geq\mathcal{L}^{>k} \, \forall \, k$ although this need not hold in general.
The core idea behind this variation on the ELBO is to sample the $k$ lowest latent variables from the conditional prior $\zb_1,\dots,\zb_l \sim p_\thetab(\zb_{\leq k}|\zb_{>k})$ and only the $L-k$ highest from the approximate posterior $\zb_{k+1},\dots,\zb_L \sim q_\phib(\zb_{>k}|\xb)$.
Importantly, this has the effect that the data likelihood $p(\xb|\zb)$ is dependent on the approximate posterior through a latent variable $\zb_{k+1}$ different from $\zb_1$ for all $k \geq 1$.
Thereby, the likelihood can be evaluated with a reconstruction from each of the latent variables $\zb_k$ of the hierarchical VAE.
Hence, we can now test how well the input $\xb$ is reconstructed from each latent variable.
The notation $\mathcal{L}^{>k}$ highlights that for latent variables $\zb_{>k}$, the bound is the regular ELBO while for the latent variables $\zb_{\leq k}$, the bound is evaluated using the (conditional) prior rather than the approximate posterior as the proposal distribution.

@@

The OOD scores considered in this research all assume that what discriminates an out-of-distribution from an in-distribution data point are semantic, high-level features. Clearly, if this is not the case and the difference instead lies in low-level statistics, the scores would likely fail. We hypothesize that a complementary bound to \cref{eq_hierarchical:biva->k}, $\mathcal{L}^{<l}$ described in \cref{sec_hierarchical:complementary-bound}, might be useful in these cases, but leave further examination to future work.

@@

\paragraph{Tasks} We follow existing literature \cite{nalisnick_deep_2019, hendrycks_deep_2019} and evaluate our method by setting up OOD detection tasks from FashionMNIST \cite{xiao_fashionmnist_2017} to MNIST \cite{lecun_gradientbased_1998} and from CIFAR10 \cite{krizhevsky_learning_2009} to SVHN \cite{netzer_reading_2011}.
For each experiment we train our model on the train split of the former dataset and test its ability to recognize the test split of the latter dataset as OOD from the test split of the former dataset.
We use the standard train/test splits for the datasets.
More details on the datasets can be found in the \cref{sec_hierarchical:datasets}.

@@

\paragraph{Models} For each OOD task, we train a simple bottom-up hierarchical VAE with $L$ stochastic layers which we will refer to as ``HVAE''.
To alleviate posterior collapse we include skip-connections that connect $\zb_i$ to $\zb_{i+2}$ for $i\in\{0, L-2\}$ and $\zb_0\equiv\xb$ in both the inference and generative models \cite{dieng_avoiding_2019} and employ the \textit{free bits} scheme with $\lambda=2$ \cite{kingma_improved_2016}.
We use weight-normalization \cite{salimans_weight_2016} on all weights and residual networks in the deterministic paths. 
A graphical representation of this model can be seen in \cref{fig_hierarchical:hvae-graphical-model}.
We use a Bernoulli output distribution for FashionMNIST/MNIST and a discretized mixture of logistics output distribution \cite{salimans_pixelcnn_2017} for CIFAR10/SVHN.
We use $L=3$ for grey-scale images and $L=4$ for natural images.
% For CIFAR/SVHN, we also train a BIVA model \cite{maaloe_biva_2019} with $L=10$ and similar configuration as used by the original paper\footnote{Code available at \url{github.com/larsmaaloee/BIVA} and \url{github.com/vlievin/biva-pytorch}}.
Full model details are in the \cref{sec_hierarchical:model-details}.

@@

The likelihoods for our trained models are in \cref{tab_hierarchical:bits-per-dim-ood} alongside baseline results for in-distribution and OOD data.
The main results of the paper on the OOD tasks can be seen along with comparisons to the baseline methods in \cref{tab_hierarchical:rocauc-ood}.
We note that for all our results, the value of the score ($\mathcal{L}^{>k}$ and $LLR^{>k}$) for the training and test splits of the in-distribution data was observed to have the same empirical distribution to within sampling error hence yielding an AUROC score of $\approx0.5$ as expected.
Results on additional commonly used datasets are found in \cref{sec_hierarchical:additional-results}.

\begin{table}[t!]
    \caption[Average bits per dimension of different datasets for generative models trained on FashionMNIST and CIFAR10.]{%
    Average bits per dimension of different datasets for models trained on FashionMNIST and CIFAR10.
    For the hierarchical models we include the $\mathcal{L}^{>k}$ bounds.
    The likelihoods of training and test splits of the in-distribution data are all cases close.
    Since we train on dynamically binarized FashionMNIST, our bits/dim are smaller than for Glow.
    As $k$ is increased for the $L^{>k}$ bound, the bound gets looser but the model eventually assigns higher likelihood to the in distribution data than to the OOD data.
    Glow refers to \textcite{kingma_glow_2018, nalisnick_deep_2019}.
    BIVA refers to our implementation of \textcite{maaloe_biva_2019}.}
    \centering
    \resizebox{0.8\columnwidth}{!}{%
    \begin{tabular}{lrrrrr}
        \toprule
         Method & Dataset & \multicolumn{4}{c}{Avg. bits/dim}\\
        %   & & $\log p(x)$ & $\mathcal{L}^{>k_1}$ & $\mathcal{L}^{>k_2}$ & $\mathcal{L}^{>k_3}$\\
          & & $\log p(x)$ & $\mathcal{L}^{>1}$ & $\mathcal{L}^{>2}$ & $\mathcal{L}^{>3}$\\
         \midrule
         \multicolumn{6}{c}{\textbf{Trained on FashionMNIST}} \\
         \midrule
         \multirow{2}{*}{Glow}
            & FashionMNIST & 2.96 & - & - & \\
            & MNIST & 1.83 & - & - & \\
         \multirow{2}{*}{HVAE (Ours)}
            & FashionMNIST & 0.420 & 0.476 & 0.579 & - \\
            & MNIST & 0.317 & 0.601 & 0.881 & - \\
         \midrule
         \multicolumn{6}{c}{\textbf{Trained on CIFAR10}} \\
         \midrule
         \multirow{2}{*}{Glow}
          & CIFAR10 & 3.46 & - & - & \\
          & SVHN & 2.39 & - & - & \\
         \multirow{2}{*}{HVAE (Ours)}
            & CIFAR10 & 3.74 & 17.8 & 54.3 & 75.7 \\  % log p(x) = 8010.03
            & SVHN & 2.62 & 10.2 & 64.0 & 93.9 \\
         \multirow{2}{*}{BIVA (Ours)}
          & CIFAR10 & 3.46 & 8.74 & 19.7 & 37.3 \\
          & SVHN & 2.35 & 6.62 & 25.1 & 59.0 \\
         \bottomrule
    \end{tabular}
    }
    \label{tab_hierarchical:bits-per-dim-ood}
    \vspace{0mm}
\end{table}

@@

\subsection{Likelihood-ratio-based OOD detection}
\begin{figure}[t]
    \centering
    \begin{subfigure}[l]{0.495\columnwidth}
        \includegraphics[width=1\columnwidth]{paper_hierarchical/roc-FashionMNIST-test-MNIST-test-ll-and-llr-IW250_sohau.pdf}
        \caption{}
        % \caption{ROC curves with AUROC score for detecting MNIST as OOD with the HVAE model trained on FashionMNIST.
        % A ROC curve is plotted for each of the $\mathcal{L}^{>k}$ bounds including the ELBO along with one for the best-performing log likelihood-ratio $LLR^{>1}$.}
        \label{fig_hierarchical:FMNIST-roc-llr}
    \end{subfigure}
    \hfill
    \begin{subfigure}[r]{0.495\columnwidth}
        \includegraphics[width=1\columnwidth]{paper_hierarchical/roc-biva-CIFAR10-SVHN-ll-and-llr_sohau.pdf}
        \caption{}
        % \caption{ROC curves with AUROC score for detecting SVHN as OOD with the BIVA model trained on CIFAR10.
        % A ROC curve is plotted for each of the $\mathcal{L}^{>k}$ bounds including the ELBO along with one for the best-performing log likelihood-ratio $LLR^{>2}$.}
        \label{fig_hierarchical:CIFAR10-roc-llr}
    \end{subfigure}
    \caption[ROC curves for out of distribution detection (MNIST/FashionMNIST and SVHN/CIFAR10).]{%
        ROC curves with AUROC score for detecting \subref{fig_hierarchical:FMNIST-roc-llr} MNIST as OOD with the HVAE model trained on FashionMNIST and \subref{fig_hierarchical:CIFAR10-roc-llr} SVHN as OOD with the BIVA model trained on CIFAR10. 
        A ROC curve is plotted for each of the $\mathcal{L}^{>k}$ bounds including the ELBO along with one for the best-performing log likelihood-ratio $LLR^{>1}$.
    }
\end{figure}
We now move to the likelihood ratio-based score.
We find that $LLR^{>k}$ separates the OOD MNIST data from in-distribution FashionMNIST to a higher degree than the likelihood estimates as can be seen by the empirical densities of the score in \cref{fig_hierarchical:FMNIST-llr}.
We note that the likelihood ratio between the ELBO and the $\mathcal{L}^{>k}$ bound provides the highest degree of separation of MNIST and FashionMNIST as measured by the AUROC$\uparrow$ for $k=1$ smaller than $L$.
This is not surprising since the value of $k$ that provides the maximal separation to the reference in-distribution dataset need not be the one for which $\mathcal{LLR}^{>k}$ is overall maximal for the OOD dataset.
We also visualize the ROC curves resulting from using the $LLR^{>k}$ score for OOD detection on both FashionMNIST/MNIST and CIFAR10/SVHN and compare it to the ROC curves resulting from the different $\mathcal{L}^{>k}$ bounds in \cref{fig_hierarchical:FMNIST-roc-llr, fig_hierarchical:CIFAR10-roc-llr}, respectively.
On both datasets we see significantly better discriminatory performance when using the $LLR^{>k}$ score.

### 2023-09-26 chapters/paper_retrospective.tex
Move table captions to top

\subsection{Population characteristics}
%
Calls to the MH-1813 were divided into training, validation, and test subsets and calls to the emergency line 1-1-2 were only used as supplementary training data (Table 1). Calls from the test year (2021) that were not associated with a diagnostic category code, which we used to evaluate call-taker performance, were separated from our primary test set, but still included to assess potential bias in this group of calls (2021 w/o category, Table 1).  The 1-1-2 training data differed from the MH-1813 data regarding age, male/female ratio, and stroke prevalence (Table 1). We therefore performed an ablation study where 1-1-2 data were not used for training to assess whether this difference negatively impacted model performance. The training, validation, and test subsets of the MH-1813 data had similar characteristics, whereas the 2021 data without diagnostic categories differed in age and sex.

@@

The classification model outperformed the call-takers (Table 21), with significant differences in all metrics (p < 0.0001). Excluding the 1-1-2 call line training data significantly degraded the model's performance (p < 0.0001), despite the domain mismatch with the MH-1813 call line test data. The performance on the 2021 calls without a diagnostic category was significantly worse than that of the test set regarding F1-score, sensitivity, false positive rate (FPR), and false omission rate (FOR) (p < 0.0001). The difference in positive predictive value (PPV) was not significant (p = 0.298).

>>>>

\subsection{Population characteristics}
%
Calls to the MH-1813 were divided into training, validation, and test subsets and calls to the emergency line 1-1-2 were only used as supplementary training data (\cref{tab_retrospective:table1-population-characteristics}). Calls from the test year (2021) that were not associated with a diagnostic category code, which we used to evaluate call-taker performance, were separated from our primary test set, but still included to assess potential bias in this group of calls (2021 w/o category, \cref{tab_retrospective:table1-population-characteristics}).  The 1-1-2 training data differed from the MH-1813 data regarding age, male/female ratio, and stroke prevalence (\cref{tab_retrospective:table1-population-characteristics}). We therefore performed an ablation study where 1-1-2 data were not used for training to assess whether this difference negatively impacted model performance. The training, validation, and test subsets of the MH-1813 data had similar characteristics, whereas the 2021 data without diagnostic categories differed in age and sex.

@@

The classification model outperformed the call-takers (\cref{tab_retrospective:table2-main-results}), with significant differences in all metrics (p < 0.0001). Excluding the 1-1-2 call line training data significantly degraded the model's performance (p < 0.0001), despite the domain mismatch with the MH-1813 call line test data. The performance on the 2021 calls without a diagnostic category was significantly worse than that of the test set regarding F1-score, sensitivity, false positive rate (FPR), and false omission rate (FOR) (p < 0.0001). The difference in positive predictive value (PPV) was not significant (p = 0.298).

### 2023-09-26 chapters/paper_review/03_approaches.tex
Move table captions to top

The masking policies used in NLP can be adapted to speech by considering a speech \edit{segment} equivalent to a token in a sentence; indeed, the masking strategy of BERT has also been used for speech pre-training~\cite{liu_mockingjay_2020}.
In the standard BERT masking policy, each token is masked independently at random. However, for speech, masking a single \edit{sample or spectrogram frame results in a largely} trivial reconstruction task \edit{since, as discussed in paragraph \ref{par:apc}, the smoothness of audio signals may encourage the model to learn to simply interpolate neighboring frames.} Therefore it is common to mask chunks of consecutive frames~\cite{liu_mockingjay_2020,jiang_further_2021}.

>>>>

The masking policies used in NLP can be adapted to speech by considering a speech \edit{segment} equivalent to a token in a sentence; indeed, the masking strategy of BERT has also been used for speech pre-training~\cite{liu_mockingjay_2020}.
In the standard BERT masking policy, each token is masked independently at random. However, for speech, masking a single \edit{sample or spectrogram frame results in a largely} trivial reconstruction task \edit{since, as discussed in \cref{par:apc}, the smoothness of audio signals may encourage the model to learn to simply interpolate neighboring frames.} Therefore it is common to mask chunks of consecutive frames~\cite{liu_mockingjay_2020,jiang_further_2021}.
